---
title: "Using the iso8601 functions"
author: "Colin Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Data interoperability is improved with the adoption of standard date and time formats. To facilitate use of the ISO 8601 standard, the Environmental Data Initiative has developed the set of `iso8601_*` functions addressing the common use cases of date time conversion, format string detection, and import to R as POSIXct and POSIXlt. This vignette provides an overview of these functions, their capabilities, and limitations.

```{r instload dataCleanr, message=FALSE, warning=FALSE}
# Install and load dataCleanr
# devtools::install_github('clnsmth/dataCleanr')
library(dataCleanr)
```

## Conversion

Converting dates represented by a single format is easy.

```{r one format, echo=TRUE, paged.print=TRUE}
# Example data
head(data_iso8601$date)

# Convert to ISO 8601
x <- iso8601_convert(x = data_iso8601$date, orders = 'mdy')
head(x)
```

Converting dates represented by two formats only requires adding the second format to the `orders` argument. If only one format is supplied to orders, then a helpful warning message is printed.

```{r two formats, echo=TRUE, paged.print=TRUE}
# Example data
data_iso8601$date_2_formats

# Convert with inadequate orders
iso8601_convert(x = data_iso8601$date_2_formats, orders = 'mdy')

# Convert with fully defined orders
iso8601_convert(x = data_iso8601$date_2_formats, orders = c('mdy', 'dmy'))
```

Time zone offsets are supplied as a two digit hour with a '+' or '-' with respect to UTC.

```{r datetimes with offset, echo=TRUE, paged.print=TRUE}
# Example data
head(data_iso8601$datetime)

# Convert to ISO 8601
x <- iso8601_convert(x = data_iso8601$datetime, orders = 'mdy HMS', tz = '-05')
head(x)
```

Converting dates represented by several formats is a bit more involved. 

```{r converting many formats, echo=TRUE, paged.print=TRUE}
# Several date and time formats representing 2012-05-15T13:29:54
data_iso8601$datetime_mess

# Convert to ISO 8601
iso8601_convert(x = data_iso8601$datetime_mess, orders = c('ymd', 'mdy', 'dmy HMS'))
```

## Datetime format string detection

Automated extraction of metadata content from data entities streamlines data archiving tasks. Retrieve the date time format specifier from a vector of ISO 8601 data created with `iso8601_convert` using `iso8601_get_format_string`.

```{r format string, echo=TRUE, paged.print=TRUE}
# Input data
data_converted <- iso8601_convert(x = data_iso8601$datetime, orders = 'mdy HMS', tz = '-05')
head(data_converted)

# Get format string
iso8601_get_format_string(x = data_converted)
```

If more than one date time format exists, `iso8601_get_format_string` returns the mode.

```{r format string mode, echo=TRUE, paged.print=TRUE}
# Input data
data_converted <- c(
  '2012-05-15T13:45:00',
  '2012-06-15T13:45:00',
  '2012-07-15T13:45:00',
  '2012-08-15T13:45',
  '2012-09-15T13:45',
  '2012-10-15T13'
)

# Get the most common format
iso8601_get_format_string(x = data_converted)
```

## Reading into R

To read ISO 8601 formatted dates and times into R as POSIXct POSIXt, use `iso8601_read`.

```{r read, echo=TRUE, paged.print=TRUE}
# Input data
data_converted <- iso8601_convert(x = data_iso8601$datetime, orders = 'mdy HMS', tz = '-05')
head(data_converted)

# Read data into R
x <- iso8601_read(x = data_converted)
head(x)
class(x)