---
title: "Functions supporting ISO 8601 datetimes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview

Create ISO 8601 character strings with `iso8601_convert`. Read ISO 8601 character strings with `iso8601_read`. Detect ISO 8601 format strings with `iso8601_get_format_string`. These functions support valid ISO8601 datetime resolutions and time zones.

- [Install and load libraries](#install-and-load-libraries)
- [Load test data](#load-test-data)
- [Convert](#convert)
- [Read](#read)
- [Format string](#format-string)

### Install and load libraries

```{r instload devtools, message=FALSE, warning=FALSE}
# Install of dataCleanr requires devtools
install.packages('devtools', repos = 'http://cran.us.r-project.org')
library(devtools)

install_github('EDIorg/dataCleanr')
library(dataCleanr)
```

* [Back to top](#overview)

### Load test data

The test data contains two columns:

* **raw** - Datetime strings in a multitude of formats
* **iso8601** - ISO 8601 formatted versions of the raw data

```{r load data, echo=TRUE, paged.print=TRUE}
data <- read.csv(
  system.file('datetimes.csv', package = 'dataCleanr'), 
  header = T,
  as.is = T
  )

data
```

* [Back to top](#overview)

### Convert

```{r convert attempt, echo=TRUE, paged.print=TRUE}
data$raw
```

* [Back to top](#overview)

### Read


* [Back to top](#overview)

### Format string


* [Back to top](#overview)
